### 2025-03-06 paper/main.tex
content, abstract

\documentclass[12pt]{article}
\usepackage[letterpaper]{geometry}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath,amsfonts}

@@

@@

\section*{\raggedright
A new formula for the area of a triangle:
Useless, but explicitly in Deep Sets and map--reduce form\footnote{%
The authors thank Alexander Novara (NYU), Ben Wandelt (Sorbonne), and Kaze Wong (JHU) for valuable discussions.}}

@@

\paragraph{Abstract:}
It is known that any permutation-invariant function of data points $r_i$
can be written in the form $\rho(\sum_i\phi(r_i))$,
where $\rho$ and $\phi$ are (probably nonlinear) functions, and the function $\phi$ might output a high-dimensional latent encoding of each point $r_i$.
This form---known in the machine-learning literature as Deep Sets---is a map--reduce form, because the $\phi$ executions can be performed asynchronously and the sums can be performed hierarchically or in any order.
The area of a triangle is a permutation-invariant function of the locations $r_i$ of the three corners $1\leq i\leq 3$.
We find, possibly for the first time, the formula for the area of a triangle, given the positions of the three corners, that is explicitly in Deep Sets form.
% It is the square-root of a permutation-invariant quartic polynomial involving the integer 368.
No insights of any kind were gained from this work.

\section{Introduction}
The area $A$ of a triangle can be written in many ways.
Here are a few:
\begin{align}
A &= \frac{1}{2}\,\text{(base)}\,\text{(height)} \label{eq:school} \\
  &= \frac{1}{2}\,|(\vec{r}_3 - \vec{r}_1) \times (\vec{r}_2 - \vec{r}_1)| \label{eq:cross} \\
  &= \frac{1}{2}\,a\,b\,\sin(C) \label{eq:sine} \\
  &= \sqrt{s\,(s-a)\,(s-b)\,(s-c)} \label{eq:Heron} ~,
\end{align}
where the first formula \eqref{eq:school} is the usual introduction to the subject,
$\vec{r}_1, \vec{r}_2, \vec{r}_3$ are three vectors pointing from an origin to the corners of the triangle,
$\times$ represents the cross-product operator (or its generalization to higher dimensions),
$a, b, c$ are side lengths of the triangle,
$C$ is the angle between the $a$ and $b$ sides,
and $s$ is the semi-perimeter $(a + b + c)/2$.
Below, we present a new formula for the area of a triangle, which we believe may be novel.

While the area of a triangle does not depend on how its sides or corners are labeled,
only the last of these formulae---Heron's formula \eqref{eq:Heron}---is explicitly permutation-invariant.

Theorem 2 of \citet{zaheer+17-Deep-Sets} states that a function $f(X)$ operating on a set $X$ is invariant to the permutation of instances in $X$ if and only if it can be decomposed in the form
\begin{equation}
    \textstyle f(X) = \rho \left( \sum_{x \in X} \phi(x) \right), \label{eq:DeepSets}
\end{equation}
for some functions $\rho$ and $\phi$.
This is proven for the cases where the elements of $X$ are from a countable universe and the size of $X$ is unlimited, and for the case where the elements of $X$ are from an uncountable universe but the size of $X$ is finite.
This result---often known by the name of the machine learning architecture derived from it, \emph{Deep Sets} \citep{zaheer+17-Deep-Sets}---is not at all obvious:
For example, even though Heron's formula \eqref{eq:Heron} is permutation-invariant, it is not actually in the form \eqref{eq:DeepSets}, if the triangle is seen as being defined by its corners (or even its sides).

A few comments about this Deep Sets form \eqref{eq:DeepSets} we'd make are the following:
\begin{itemize}
    \item The function $\phi$ is universal to all the elements $x$; its form won't depend on label $i$ or on any properties of each point, nor any properties of the overall collection of points (the \emph{point cloud}, as it were).
    \item The point properties $r_i$ and the output of the function $\phi$ might both be very large.
    That is, each point $r_i$ might be a vector or even an ordered list of vectors and tensors and images and so on, and the output of the function $\phi$ might also be very large and complicated.
    The output of the function $\phi$ is some kind of informative, latent encoding of the point $r_i$.
    \item If a function can be written in Deep-Sets form \eqref{eq:DeepSets}, then that function can be executed in a map--reduce framework, in which the function $\phi$ is executed independently and asynchronously on each data point (maybe even in the data center) and the sums are performed hierarchically or in any order.
    That means that for large problems executed on large data sets in large data centers, problems in this form \eqref{eq:DeepSets} can often be performed in wall-clock time that scales as $\log N$.
\end{itemize}

@@

Perhaps non-obviously, for a monomial to be in Deep Sets form, it must consist only of products of sums over the three points.
It cannot contain any differences, or any sums over pairs of points (or edges).
Thus $(\sum_{i=1}^2 x_i^2)\,(\sum_{i=1}^3 x_i\,y_i)$ is a permitted monomial, but $(\sum_{i=1}^3\sum_{j=1}^3 x_i^2\,y_j^2)$ is not a permitted monomial.
These monomials are both correct from a units perspective, but the second is not in Deep Sets form, because it contains products of position coordinates from different points. 
In order to generate all the possible invariant monomials, we first note that they are made of products of factors of the form $F_{m,n} = \sum_i x_i^m y_i^n$.
The invariant monomials of degree $d$ are then all products $\prod_i F_{m_i, n_i}$ such that $\sum_i m_i + n_i = d$.

% HOGG: How to construct all the invariant polynomials.

% HAINJE: Given data of form $(x, y)$, invariant factors are $F_{m,n} = \sum_i x_i^m y_i^n$, and the invariant monomials are any products of these factors $\prod_{i} F_{m_i,n_i}$. Do we want/need to be more specific than that?

@@

Our result---and only contribution---is that the area $A$ of a triangle is given by
% \begin{align}
% A^2 = &
% -\frac{ 54}{368}\,(\sum_{i=1}^3 x_i^2\,y_i^2) 
% -\frac{258}{368}\,(\sum_{i=1}^3 x_i\,y_i)^2 \nonumber \\ &
% +\frac{285}{368}\,(\sum_{i=1}^3 x_i^2)\,(\sum_{i=1}^3 y_i^2) 
% +\frac{ 36}{368}\,(\sum_{i=1}^3 y_i)\,(\sum_{i=1}^3 x_i^2\,y_i) \nonumber \\ &
% +\frac{ 36}{368}\,(\sum_{i=1}^3 x_i)\,(\sum_{i=1}^3 x_i\,y_i^2) 
% -\frac{101}{368}\,(\sum_{i=1}^3 y_i)^2\,(\sum_{i=1}^3 x_i^2) \nonumber \\ &
% +\frac{148}{368}\,(\sum_{i=1}^3 x_i)\,(\sum_{i=1}^3 y_i)\,(\sum_{i=1}^3 x_i\,y_i) 
% -\frac{101}{368}\,(\sum_{i=1}^3 x_i)^2\,(\sum_{i=1}^3 y_i^2) \nonumber \\ &
% +\frac{  9}{368}\,(\sum_{i=1}^3 x_i)^2\,(\sum_{i=1}^3 y_i)^2 ~, \label{eq:result}
% \end{align}
\begin{align}
    A^2 = & \
    \frac{2}{4} \, (\sum_{i=1}^{3} x_i ) \, (\sum_{i=1}^{3} y_i ) \, (\sum_{i=1}^{3} x_i \, y_i )
    \nonumber \\ &
    + \frac{3}{4} \, (\sum_{i=1}^{3} x_i^2 ) \, (\sum_{i=1}^{3} y_i^2 )
    - \frac{3}{4} \, (\sum_{i=1}^{3} x_i \, y_i )^2.
    \nonumber \\ &
    - \, \frac{1}{4} \, (\sum_{i=1}^{3} x_i )^2 \, (\sum_{i=1}^{3} y_i^2 )
    - \, \frac{1}{4} \, (\sum_{i=1}^{3} y_i )^2 \, (\sum_{i=1}^{3} x_i^2 )
     ~, \label{eq:result}
\end{align}
where the two-dimensional position of each of the three corners $i$ is $r_i \equiv (x_i, y_i)$.
This answer is in Deep Sets form:
Every expression in parentheses in \eqref{eq:result} is a pure sum over all three points and can be written in the form of $F_{mn}$ for some choice of $m$ and $n$.
There are no sums over differences, and there is no ordering of the points.
Being very explicit, this can be written in the Deep Sets form \eqref{eq:DeepSets} if we define $\rho$ and $\phi$ as follows:
\begin{gather}
    \phi(\vec{r}_i) = (
        x_i, \,
        y_i, \,
        x_i^2, \,
        x_i \, y_i, \,
        y_i^2
    ), \\
    {\textstyle \sum_i} \phi(\vec{r}_i) = (
        F_{10}, \,
        F_{01}, \,
        F_{20}, \,
        F_{11}, \,
        F_{02}
    ), \\
    \rho \left( {\textstyle \sum_i} \phi(\vec{r}_i) \right)
    = \tfrac{2}{4} \, F_{10} \, F_{01} \, F_{11}
    + \tfrac{3}{4} \, F_{20} \, F_{02}
    - \tfrac{3}{4} \, F_{11}^2
    - \tfrac{1}{4} \, F_{10}^2 \, F_{02}
    - \tfrac{1}{4} \, F_{01}^2 \, F_{20}.
\end{gather}

@@

But note that it isn't obviously invariant to arbitrary rotations.
And it isn't obviously invariant to translations.
Formulae \eqref{eq:cross}, \eqref{eq:sine}, and \eqref{eq:Heron} are obviously rotation-invariant and translation-invariant.

If the triangle is embedded in a higher-dimensional space with the positions of the three vertices given by $\vec{r}_i \in \mathbb{R}^{d}$ for $d \geq 2$, we note that the area of the triangle can be determined by summing in quadrature the areas of the triangles made by all projections to two axes. For example, in three dimensions, the area of the triangle is given by
\begin{equation}
    A_{\rm 3D}^2 =
    A_{12}^2 + A_{13}^2 + A_{23}^2,
\end{equation}
and, in four dimensions, it is given by
\begin{equation}
    A_{\rm 4D}^2 =
    A_{12}^2 + A_{13}^2 + A_{14}^2
    + A_{23}^2 + A_{24}^2
    + A_{34}^2,
\end{equation}
where we have introduced $A_{jk}$ the area of the triangle formed by just the $j^{\rm th}$ and $k^{\rm th}$ components of the position vectors.

Our method also extends to the hypervolume of a $d$-simplex. Our findings are simplified by the realization that our result is a polynomial with \emph{integer} coefficients for the squared area of a parallelogram. Expecting this to generalize, we note that the hypervolume of the $d$-simplex ($S$) formed by vectors $\{ \vec{v}_{0}, \vec{v}_{1}, \cdots, \vec{v}_{d} \}$ is related to the hypervolume of the $d$-parallelotope ($P$) spanned by the vectors $\{ \vec{v}_{1} - \vec{v}_0, \cdots, \vec{v}_d - \vec{v}_0 \}$ by
\begin{equation}
    {\rm volume}(P)
    = (d!) \, {\rm volume}(S)
\end{equation}
The squared hypervolume of a $d$-parallelotope is given by the determinant of the Gram matrix of its constituent vectors, e.g.,
\begin{equation}
    {\rm volume}(P)^2
    = \det G, \quad
    G_{ij} \equiv (\vec{v}_i - \vec{v}_0) \cdot (\vec{v}_j - \vec{v}_0).
\end{equation}
In the two-dimensional case, our formula \eqref{eq:result} gives an integer polynomial for the area of the parallelogram, equal to $4 A^2$ for $A$ the area of the triangle. To find the hypervolumes of higher-dimensional simplexes, we modify our symbolic regression to seek integer polynomials for the squared volume of the related parallelotope.

In three dimensions, four vectors form the vertices of a simplex. We find an analogous Deep Sets-form formula for the volume of the simplex to be
\begin{align}
(6 V)^2 = &
+8 \, F_{1100} \, F_{1010} \, F_{0110}
+4 \, F_{200} \, F_{020} \, F_{0020}
\nonumber \\ &
-4 \, F_{200} \, F_{011}^2
-4 \, F_{020} \, F_{101}^2
-4 \, F_{002} \, F_{110}^2
\nonumber \\ &
+2 \, F_{200} \, F_{011} \, F_{010} \, F_{001}
+2 \, F_{020} \, F_{101} \, F_{100} \, F_{001}
+2 \, F_{002} \, F_{110} \, F_{100} \, F_{010}
\nonumber \\ &
-2 \, F_{110} \, F_{101} \, F_{010} \, F_{001}
-2 \, F_{110} \, F_{011} \, F_{100} \, F_{001}
-2 \, F_{101} \, F_{011} \, F_{100} \, F_{010}
\nonumber \\ &
+1 \, F_{110}^2 \, F_{001}^2
+1 \, F_{101}^2 \, F_{010}^2
+1 \, F_{011}^2 \, F_{100}^2
\nonumber \\ &
-1 \, F_{200} \, F_{020} \, F_{001}^2
-1 \, F_{200} \, F_{002} \, F_{010}^2
-1 \, F_{020} \, F_{002} \, F_{100}^2
\label{eq:simplex}
\end{align}

HOGG SAY: CH: EXPAND THIS INTO AN OPEN QUESTIONS LIST:
\begin{itemize}
\item CH should answer this question: What's the generalization to $d$ dimensions? That is, what if the three points are each in d dimensions?
\item questionable: What happens if we try to do $A$ as a quadratic instead of $A^2$ as a quartic?
\item What happens if you hold the formula fixed, but use more than $N=3$ points?
The formula still returns a squared area.
That's the area of what?
\item What about areas of quadrangles or larger polygons?
\item What about volumes and hypervolumes of larger objects?
\end{itemize}

\newpage

\newpage

Define the ``invariant factor'' in $d$ dimensions
\begin{equation}
    F_{m_1 \, m_2 \, \cdots} \equiv \sum_{1=1}^{3} \prod_{k=1}^{d} (\vec{r}_{i})_{k}^{m_k}.
\end{equation}
This is the sum over all three particles of the product of the $d$ components of each $\vec{r}_i$, each component raised to a power. For example, in 2D, $d=2$ and this expands to
\begin{equation}
    F_{\ell m} = \sum_{i=1}^{3} x_i^\ell \, y_i^m.
\end{equation}
In 3D, $d=3$ and this gives
\begin{equation}
    F_{\ell m n} = \sum_{i=1}^{3} x_i^\ell \, y_i^m \, z_i^n.
\end{equation}
Then, the area formula found above can be extended to more dimensions with the following formulae.

\begin{equation}
\begin{aligned}
V^2 = &
+8 F_{110} F_{101} F_{011}
+4 F_{200} F_{020} F_{002}
\\ &
-4 F_{200} F_{011}^2
-4 F_{020} F_{101}^2
-4 F_{002} F_{110}^2
\\ &
+2 F_{200} F_{011} F_{010} F_{001}
+2 F_{020} F_{101} F_{100} F_{001}
+2 F_{002} F_{110} F_{100} F_{010}
\\ &
-2 F_{110} F_{101} F_{010} F_{001}
-2 F_{110} F_{011} F_{100} F_{001}
-2 F_{101} F_{011} F_{100} F_{010}
\\ &
+1 F_{110}^2 F_{001}^2
+1 F_{101}^2 F_{010}^2
+1 F_{011}^2 F_{100}^2
\\ &
-1 F_{200} F_{020} F_{001}^2
-1 F_{200} F_{002} F_{010}^2
-1 F_{020} F_{002} F_{100}^2
\end{aligned}
\end{equation}

>>>>

\documentclass[12pt]{article}
\usepackage[letterpaper]{geometry}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath,amsfonts,tikz}
\usetikzlibrary{angles,calc,quotes}

@@

\usepackage{xcolor}
\newcommand{\CH}[1]{{\color{blue} (CH: #1)}}

@@

\section*{\raggedright
A new formula for the area of a triangle:
Useless, but explicitly in Deep Sets form\footnote{%
The authors thank Alexander Novara (NYU), Ben Wandelt (Sorbonne), and Kaze Wong (JHU) for valuable discussions.}}

@@

\paragraph{Abstract:}
It is known that any permutation-invariant function of data points $r_i$
can be written in the form $\rho(\sum_i\phi(r_i))$,
where $\rho$ and $\phi$ are (probably nonlinear) functions, and the function $\phi$ might output a high-dimensional latent encoding of each point $r_i$.
This form---known in the machine-learning literature as Deep Sets---is also a map--reduce, because the $\phi$ executions can be performed asynchronously and the sums can be performed hierarchically or in any order.
The area of a triangle is a permutation-invariant function of the locations $r_i$ of the three corners $1\leq i\leq 3$.
We find, possibly for the first time, the formula for the area of a triangle, given the positions of the three corners, that is explicitly in Deep Sets form.
This project was motivated by questions about the fundamental computational complexity of n-point statistics in cosmology; that said, no insights of any kind were gained from these results.

\section{Introduction}

\begin{figure}
    \centering
    \begin{tikzpicture}
    \coordinate (O) at (0,0);
    \coordinate (A) at (3,4);
    \coordinate (B) at (5,1);

    \pgfmathsetmacro{\a}{sqrt(3*3 + 4*4)}
    \pgfmathsetmacro{\b}{sqrt(5*5 + 1*1)}
    \pgfmathsetmacro{\c}{sqrt(2*2 + 3*3)}
    \pgfmathsetmacro{\s}{(\a+\b+\c)/2}
    \pgfmathsetmacro{\h}{2*sqrt(\s*(\s-\a)*(\s-\b)*(\s-\c))/\b}

    \pgfmathsetmacro{\DeltaX}{5 - 0}  % Bx - Ox
    \pgfmathsetmacro{\DeltaY}{1 - 0}  % By - Oy
    \pgfmathsetmacro{\len}{sqrt(\DeltaX*\DeltaX + \DeltaY*\DeltaY)}

    \def\bLen{0.5mm}
    \def\hLen{0.3mm}
    \pgfmathsetmacro{\bdx}{-\DeltaY/\len * \bLen}
    \pgfmathsetmacro{\bdy}{\DeltaX/\len * \bLen}
    \pgfmathsetmacro{\hdx}{\DeltaX/\len * \hLen}
    \pgfmathsetmacro{\hdy}{\DeltaY/\len * \hLen}
    \pgfmathsetmacro{\hx}{-\DeltaY/\len * \h}
    \pgfmathsetmacro{\hy}{\DeltaX/\len * \h}

    \coordinate (BaseO) at ($(O) - (\bdx,\bdy)$);
    \coordinate (BaseB) at ($(B) - (\bdx,\bdy)$);
    \coordinate (HeightO) at ($(B) + (\hdx,\hdy)$);
    \coordinate (HeightA) at ($(B) + (\hdx,\hdy) + (\hx,\hy)$);

    % mark points
    \fill (O) circle (1.2pt) node[below right]{$(x_1, y_1)$};
    \fill (A) circle (1.2pt) node[above, yshift=+1mm]{$(x_2, y_2)$};
    \fill (B) circle (1.2pt) node[below right]{$(x_3, y_3)$};

    % draw vectors
    \def\pad{1.5mm}
    \draw[->, line width=0.4mm]
        ($(O)!\pad!(A)$) -- ($(A)!\pad!(O)$)
        node[midway, left, xshift=-1mm, yshift=+1mm] {$\vec{a}$};
    \draw[->, line width=0.4mm]
        ($(O)!\pad!(B)$) -- ($(B)!\pad!(O)$)
        node[midway, below, xshift=+1mm, yshift=-0.5mm] {$\vec{b}$};
    \draw[->, line width=0.4mm]
        ($(B)!\pad!(A)$) -- ($(A)!\pad!(B)$)
        node[midway, right, xshift=+1mm, yshift=+1mm] {$\vec{c}$};

    % draw arc for angle
    \draw pic["$C$", draw=black, angle radius=12mm, angle eccentricity=1.3] {angle=B--O--A};

    % draw base, height guide lines
    \draw ($(O)!\pad!(BaseO)$) -- ($(BaseO)$);
    \draw ($(B)!\pad!(BaseB)$) -- ($(BaseB)$);
    \draw ($(B)!\pad!(HeightO)$) -- ($(HeightO)$);
    \draw ($(A)!\pad!(HeightA)$) -- ($(HeightA)$);

    % draw base, height size lines
    \draw[<->]
        ($(BaseO)!\pad!(O)!\pad!(BaseB)$) -- ($(BaseB)!\pad!(B)!\pad!(BaseO)$)
        node[midway, below, xshift=+2mm, yshift=-1mm] {base};
    \draw[<->]
        ($(HeightO)!\pad!(O)!\pad!(HeightA)$) -- ($(HeightA)!\pad!(A)!\pad!(HeightO)$)
        node[midway, right, xshift=+1mm, yshift=+1mm] {height};

    \end{tikzpicture}
    \caption{A triangle.}
    \label{fig:triangle}
\end{figure}

In Figure~\ref{fig:triangle}, we present a triangle.
The area $\Delta$ of this triangle can be written in many ways.
Here are a few:
\begin{align}
\Delta 
&= \frac{1}{2} \, \text{(base)} \, \text{(height)} \label{eq:school} \\
&= \frac{1}{2}\, \| \vec{a} \times \vec{b} \| \label{eq:cross} \\
&= \frac{1}{2}\,a\,b\,\sin(C) \label{eq:sine} \\
&= \sqrt{s\,(s-a)\,(s-b)\,(s-c)} \label{eq:Heron} ~,
\end{align}
where the first formula \eqref{eq:school} is the usual introduction to the subject,
$\times$ represents the cross-product operator,
$a = \| \vec{a} \|$, $b = \| \vec{b} \|$, $c = \| \vec{c} \|$ are side lengths of the triangle,
% $C$ is the angle between $\vec{a}$ and $\vec{b}$,
and $s$ is the semi-perimeter $(a + b + c)/2$.
For a rather exhaustive list of 110 unique formulae, see \citet{baker1885a,baker1885b}.

While the area of the triangle does not depend on how its sides or corners are labeled,
only the last of these formulae---Heron's formula \eqref{eq:Heron}---is explicitly permutation-invariant.

Permutation invariance is an important symmetry in a number of fields for problems defined on \textit{sets}.
In machine learning, one area of relevance is deep learning on point clouds (e.g., sets of three-dimensional position vectors) \citep{guo2020pointclouds}, which is particularly useful for problems such as 3D shape classification and object detection.
In physics, many fundamental problems are invariant to permutation.
For example, the gravitational/electric potential is invariant to the labeling of point masses/charges \CH{citation needed?}.
Further, permutation invariance lies at the heart of all modern cosmological statistics, such as the power spectrum and the bispectrum \CH{cite}.

There is a simple and powerful result which is useful for constructing and characterizing permutation-invariant functions.
Given a function $f(r_1, \cdots, r_N)$, Theorem 2 of \citet{zaheer+17deepsets} states that $f$ is invariant to permutation of its arguments if and only if it can be decomposed in the form
\begin{equation}
    f(r_1, \cdots, r_N) = \rho \left( {\textstyle \sum_{i=1}^{N}} \, \phi(r_i) \right), \label{eq:DeepSets}
\end{equation}
for some functions $\rho$ and $\phi$.

This result---often known by the name of the deep learning architecture derived from it, \emph{Deep Sets} \citep{zaheer+17deepsets}---is not at all obvious:
For example, even though Heron's formula \eqref{eq:Heron} is permutation-invariant, it is not actually in the form of \eqref{eq:DeepSets}. Consider instead the formula
\begin{equation}
    \Delta = \frac{1}{4} \sqrt{
        \left( a^2 + b^2 + c^2 \right)^2
        - 2 \left( a^4 + b^4 + c^4 \right)
    }.
\end{equation}
It is simple to show that this formula is exactly equivalent to Heron's formula when expanded, but it is now precisely in Deep Sets form. Explicitly,
\begin{equation}
    \phi(x) = (x^2, x^4), \quad
    \vec{u} = {\textstyle \sum_{x \in \{ a, b, c \}}} \, \phi(x), \quad
    \Delta = \rho(\vec{u}) = \frac{1}{4} \sqrt{ u_1^2 - 2 \, u_2 }.
\end{equation}

A few comments about this Deep Sets form \eqref{eq:DeepSets} we'd make are the following:
\begin{itemize}
    \item The function $\phi$ is universal to all the elements $r_i$; its form won't depend on label $i$ or on any properties of each point, nor any properties of the overall collection of points (the \emph{point cloud}, as it were).
    \item The point properties $r_i$ and the output of the function $\phi$ might both be very large.
    That is, each point $r_i$ might be a vector or even an ordered list of vectors and tensors and images and so on, and the output of the function $\phi$ might also be very large and complicated.
    The output of the function $\phi$ is some kind of informative, latent encoding of the point $r_i$.
    \item If a function can be written in Deep Sets form \eqref{eq:DeepSets}, then that function can be executed in a map--reduce framework, in which the function $\phi$ is executed independently and asynchronously on each data point (maybe even in the data center) and the sums are performed hierarchically or in any order.
    That means that for large problems executed on large data sets in large data centers, problems in this form \eqref{eq:DeepSets} can often be performed in wall-clock time that scales as $\log N$.
\end{itemize}

@@

Perhaps non-obviously, for a monomial to be in Deep Sets form, it must consist only of products of sums over the three points.
It cannot contain any differences, or any sums over pairs of points (or edges).
Thus $(\sum_{i=1}^2 x_i^2)\,(\sum_{i=1}^3 x_i\,y_i)$ is a permitted monomial, but $(\sum_{i=1}^3\sum_{j=1}^3 x_i^2\,y_j^2)$ is not a permitted monomial.
These monomials are both correct from a units perspective, but the second is not in Deep Sets form, because it contains products of position coordinates from different points. 
All the possible monomials are made of permutation-invariant factors of the form 
\begin{equation}
F_{mn} = \sum_{i=1}^{3} x_i^m \, y_i^n.
\end{equation}
The invariant monomials of degree $d$ are then all products $\prod_j F_{m_j n_j}$ such that $\sum_j m_j + n_j = d$.

@@

Our result is that the area $A$ of a triangle is given by
% \begin{align}
% A^2 = &
% -\frac{ 54}{368}\,(\sum_{i=1}^3 x_i^2\,y_i^2) 
% -\frac{258}{368}\,(\sum_{i=1}^3 x_i\,y_i)^2 \nonumber \\ &
% +\frac{285}{368}\,(\sum_{i=1}^3 x_i^2)\,(\sum_{i=1}^3 y_i^2) 
% +\frac{ 36}{368}\,(\sum_{i=1}^3 y_i)\,(\sum_{i=1}^3 x_i^2\,y_i) \nonumber \\ &
% +\frac{ 36}{368}\,(\sum_{i=1}^3 x_i)\,(\sum_{i=1}^3 x_i\,y_i^2) 
% -\frac{101}{368}\,(\sum_{i=1}^3 y_i)^2\,(\sum_{i=1}^3 x_i^2) \nonumber \\ &
% +\frac{148}{368}\,(\sum_{i=1}^3 x_i)\,(\sum_{i=1}^3 y_i)\,(\sum_{i=1}^3 x_i\,y_i) 
% -\frac{101}{368}\,(\sum_{i=1}^3 x_i)^2\,(\sum_{i=1}^3 y_i^2) \nonumber \\ &
% +\frac{  9}{368}\,(\sum_{i=1}^3 x_i)^2\,(\sum_{i=1}^3 y_i)^2 ~, \label{eq:result}
% \end{align}
\begin{align}
    A^2 = & \
    \frac{2}{4} \, (\sum_{i=1}^{3} x_i \, y_i ) \, (\sum_{i=1}^{3} x_i ) \, (\sum_{i=1}^{3} y_i )
    \nonumber \\ &
    + \frac{3}{4} \, (\sum_{i=1}^{3} x_i^2 ) \, (\sum_{i=1}^{3} y_i^2 )
    - \frac{3}{4} \, (\sum_{i=1}^{3} x_i \, y_i )^2.
    \nonumber \\ &
    - \, \frac{1}{4} \, (\sum_{i=1}^{3} y_i^2 ) \, (\sum_{i=1}^{3} x_i )^2
    - \, \frac{1}{4} \, (\sum_{i=1}^{3} x_i^2 ) \, (\sum_{i=1}^{3} y_i )^2
     ~, \label{eq:result}
\end{align}
where the two-dimensional position of each of the three corners $i$ is $r_i \equiv (x_i, y_i)$.
This answer is in Deep Sets form:
Every expression in parentheses in \eqref{eq:result} is a pure sum over all three points and can be written in the form of $F_{mn}$ for some choice of $m$ and $n$.
There are no sums over differences, and there is no ordering of the points.
Being very explicit, this is written in the Deep Sets form \eqref{eq:DeepSets} for $\rho$ and $\phi$ defined as follows:
\begin{gather}
    \phi(\vec{r}_i) = (
        x_i, \,
        y_i, \,
        x_i^2, \,
        x_i \, y_i, \,
        y_i^2
    ), \\
    {\textstyle \sum_i} \phi(\vec{r}_i) = (
        F_{10}, \,
        F_{01}, \,
        F_{20}, \,
        F_{11}, \,
        F_{02}
    ), \\
    \rho \left( {\textstyle \sum_i} \phi(\vec{r}_i) \right)
    = \tfrac{2}{4} \, F_{11} \, F_{10} \, F_{01}
    + \tfrac{3}{4} \, F_{20} \, F_{02}
    - \tfrac{3}{4} \, F_{11}^2
    - \tfrac{1}{4} \, F_{02} \, F_{10}^2
    - \tfrac{1}{4} \, F_{20} \, F_{01}^2.
\end{gather}

@@

% But note that it isn't obviously invariant to arbitrary rotations.
% And it isn't obviously invariant to translations.
% Formulae \eqref{eq:cross}, \eqref{eq:sine}, and \eqref{eq:Heron} are obviously rotation-invariant and translation-invariant.

Note that a triangle defined by three points has the one-half the area of the parallelogram spanned by the vectors $\vec{r}_2 - \vec{r}_1$ and $\vec{r}_3 - \vec{r}_1$. Multiplying both sides by 4 then reveals that \eqref{eq:result} is an integer polynomial of invariant factors for the area of the parallelogram.

If the triangle is embedded in a higher-dimensional space with the positions of the three vertices given by $\vec{r}_i \in \mathbb{R}^{d}$ for $d \geq 2$, the area of the triangle can be determined by summing in quadrature the areas of the triangles made by all projections to two axes. For example, in three dimensions, the area of the triangle is given by
\begin{equation}
    A_{\rm 3D}^2 =
    A_{12}^2 + A_{13}^2 + A_{23}^2,
\end{equation}
and, in four dimensions, it is given by
\begin{equation}
    A_{\rm 4D}^2 =
    A_{12}^2 + A_{13}^2 + A_{14}^2
    + A_{23}^2 + A_{24}^2
    + A_{34}^2,
\end{equation}
where we have introduced $A_{jk}$, the area of the triangle formed by just the $j^{\rm th}$ and $k^{\rm th}$ components of the position vectors. By using \eqref{eq:result} for each of the two-dimensional sub-areas, each can be computed in a permutation-invariant way and the entire result is thus permutation invariant (and in Deep Sets form).

Our method also extends to the volume of a simplex. In three dimensions, four vectors $\{ \vec{r}_1, \vec{r}_2, \vec{r}_3, \vec{r}_4 \}$ determine the vertices of a simplex. 
Like the area of a triangle is related to the area of a parallelogram, the volume of a simplex is one-sixth the volume of the parallelepiped spanned by the vectors $\{ \vec{r}_2 - \vec{r}_1, \vec{r}_3 - \vec{r}_1, \vec{r}_4 - \vec{r}_1 \}$. Hoping for an integer solution once again, we modify our symbolic regression to search for polynomials of the invariant factors which yield the volume of the parallelepiped. This gives an analogous Deep Sets formula for the volume of a simplex to be
\begin{align}
(6 V)^2 = &
+8 \, F_{110} \, F_{101} \, F_{011}
+4 \, F_{200} \, F_{020} \, F_{002}
\nonumber \\ &
-4 \, F_{200} \, F_{011}^2
-4 \, F_{020} \, F_{101}^2
-4 \, F_{002} \, F_{110}^2
\nonumber \\ &
+2 \, F_{200} \, F_{011} \, F_{010} \, F_{001}
+2 \, F_{020} \, F_{101} \, F_{100} \, F_{001}
+2 \, F_{002} \, F_{110} \, F_{100} \, F_{010}
\nonumber \\ &
-2 \, F_{110} \, F_{101} \, F_{010} \, F_{001}
-2 \, F_{110} \, F_{011} \, F_{100} \, F_{001}
-2 \, F_{101} \, F_{011} \, F_{100} \, F_{010}
\nonumber \\ &
+1 \, F_{110}^2 \, F_{001}^2
+1 \, F_{101}^2 \, F_{010}^2
+1 \, F_{011}^2 \, F_{100}^2
\nonumber \\ &
-1 \, F_{200} \, F_{020} \, F_{001}^2
-1 \, F_{200} \, F_{002} \, F_{010}^2
-1 \, F_{020} \, F_{002} \, F_{100}^2 \, .
\label{eq:simplex}
\end{align}

% HOGG SAY: CH: EXPAND THIS INTO AN OPEN QUESTIONS LIST:
% \begin{itemize}
% \item questionable: What happens if we try to do $A$ as a quadratic instead of $A^2$ as a quartic?
% \item What happens if you hold the formula fixed, but use more than $N=3$ points?
% The formula still returns a squared area.
% That's the area of what?
% \item What about areas of quadrangles or larger polygons?
% \item What about volumes and hypervolumes of larger objects?
% \end{itemize}

We leave the reader with the following open questions:
\begin{itemize}
    \item It is trivial to take the form of \eqref{eq:result} and evaluate the sums for more than three vectors. Does the result have any meaning?

    \item Our formula for the volume of a simplex \eqref{eq:simplex} bears many similarities to that for the area of triangle \eqref{eq:result}. We take this to imply that a similar formula should exist for the hypervolume of a $d$-simplex. What is this result?

    \item In both the triangle area and simplex volume formulae, the only invariant factors used are of degree $2$, and the factors never appear with exponents larger than $2$. Does this generalize? Can it be shown that higher order factors aren't needed?
\end{itemize}
